Data Specialization for a Staged Lambda Calculus

### 2015-02-26 paper/intro.tex
minor

Such computations, where we apply a multivariate function to arguments
where one of the argument remains the same are common. In fact, many
research problems aim to improve the efficiency of such functions.  For
example, algorithm design techniques such as dynamic programming
identify the computation in $f$ that depend on $x$, and try to ensure that
such computations are performed no more than once.  Similarly, crucial
compiler optimization techniques such as loop hoisting and common
subexpression elimination identify common computations across all
invocations of $f$ and perform them once and for all at the beginning
of the computation. The whole area of incremental computation aims at
adapting a computation to small changes, which may be applicable to
this problem if the $y_i$'s are sufficiently similar.

The common theme in all such design and optimization techniques is to
achieve what in the 1980s J{\o}rring and Scherlis called {\em
frequency reduction} or {\em precomputation}~\cite{JS86-staging}.
The idea behind frequency reduction is to identify computations that
are performed multiple times and hoist them so that they can be
performed once and used later as needed.  The idea behind
precomputation is to identify computations that can be performed
earlier---for example, at compile time if their inputs are known
statically---and perform them at that earlier time.

@@

Partial evaluation achieves precomputation by distinguishing between 
{\em static} and {\em dynamic} stages, and performing all static
computations first, followed by dynamic ones.  For example,
suppose that we wish to compute $F(x, y_0 \ldots y_{m-1})$ for a
statically known $x = [~]$.  We can derive a specialization of $f$ at
$[~]$, written $f_{[~]}$, such that $f(x,y) = f_{[~]}(y)$ and specialize $F$
for $x=[~]$ as
\begin{lstlisting}
fun @$F_{[~]}(y_0, y_1 \ldots y_{m-1})$@ = 
  @$(f_{[~]}(y_0);  f_{0}(y_1)  \ldots f_{0}(y_{m-1}))$@.
\end{lstlisting}
%
Such a partial evaluation typically proceed by performing a
binding-time analysis to determine the static and dynamic parts of $f$
and $F$ and specializes them based on the results of the binding-time
analysis.

@@

The third and final approach, called {\em pass separation} or {\em
  data specialization}, assumes much less information: instead of
requiring the static data, it requires just the knowledge of which
computations are static and dynamic and uses this knowledge to
generate two stage computations. 
%
Applied to our example, pass separation would {\em split} the function
$f$ into $f_s$ (static) and $f_d$ (dynamic) such that $f(x,y) =
f_d(f_s(x),y)$.  We can then rewrite our example as
%
\begin{lstlisting}
fun @$F(x, y_0, y_1 \ldots y_{m-1})$@ = 
let @$fx = f_s(x)$@
in @$(f_d(fx, y_0);  f_d(fx, y_1)  \ldots f_d(fx, y_{m-1}))$@.
\end{lstlisting}
We refer to such a program as {\em stratified} program.

The key difference with partial evaluated program and the stratified
program is that the latter has lost no generality: it works with all
arguments, whereas the partially evaluated program is specialized for
a particular value of $x$ ($x = [~]$ in the example).  Thus, the pass
separation can be applied without leading to any loss of generality.
In fact, this is quite natural---in the terminology of
J{\o}rring and Scherlis, pass separation is a frequency reduction,
whereas as partial evaluation and metaprogramming are precomputation
techniques. Consequently, if desired, partial evaluation can be
orthogonally applied to a stratified program.

% Let's assume for concreteness that $f(x,y)$ takes time in the size of
% the first argument.  The function $F(x, y_0 \ldots y_{m-1}$ then takes
% time $O(n \cdot m)$ where $n$ is the size of $x$. Assume further that
% in the stratified code the static function $f_s(x)$ takes $O(n)$ time
% and the $f_d(fsx, \cdot)$ function takes $O(\log{n})$ time.  With
% these run times, the stratified code has complexity $O(n + m\log{n})$
% instead of $O(n \cdot m)$, a significant asymptotic improvement.  (In
% this paper, we will achieve such a reduction in complexity
% automatically.)

But what is the benefit of pass separation? The big benefit is
efficiency. By stratifying the program, pass separation factors out
frequently used the parts of computation, computes them first, and
then reuse them later instead of computing them again.  Since it can
be applied statically with no loss of generality, pass separation can
significantly improve performance.  In fact, in computer graphics,
program are typically manually stratified both for improved
efficiency. Such stratification is in fact required by the execution
model of some modern graphics architectures.

While highly desirable, manual stratification is difficult and highly
nonmodular: program invariants are scattered over the code written in
multiple strata and because modification to any one stratum requires
modifying all others to match.  Ideally, we wish to write just one
program and stratify it into multiple strata automatically.  But such
automatic stratification turns out to be challenging: the first

Such automatic stratification turns out to be challenging: the first
successful approach that we know of is by Knoblock and
Ruf\,\cite{knoblock96}, which, like our work, was inspired by computer
graphics applications.  While practically effective, Knoblock and Ruf
consider only a very simple language and are not able to stratify
recursive programs.  Recent
efforts\,\cite{Proudfoot:2001,Foley:2011,He:2014} ....  but still
cannot stratify recursive programs.
%
In fact back in 80,s Jorring and Sherlis was able to stratify
reasonably complex programs but only manually and note that
``... partial evaluation technique is readily mechanized, while
staging transformation approach will elude full automation for some
time.''

In this paper we make significant progress on the problem of automatic
stratification of higher-order functional programs.  As our starting
point, we take the modal language \lang\ with explicit staging in
the style of Davies~\cite{davies96}, where the ``circle'' modality
denotes computation in a later stage, and present a staged operational
semantics similar for \lang\ (\ref{sec:stagedsemantics}).  We then
present a splitting algorithm that stratifies a \lang\ program into
a two mono-staged programs, one for the first stage and for the second
stage (\ref{sec:splitting}).  Since each program is monostaged, they
are expressed in conventional functional languages.

@@

We start in \ref{sec:example} with an extended example of splitting
\texttt{quickselect}.
In \ref{sec:semantics}, we describe our staged language \lang, including its
type system and operational semantics.
In \ref{sec:splitting,sec:implementation}, we describe our stage-splitting
algorithm for performing data specialization.
Finally, in \ref{sec:examples}, we show how our stage-splitting algorithm
transforms a variety of other programs.

>>>>

Such computations, where we apply a multivariate function to arguments
where one of the argument remains the same are common.  For example,
many algorithm-design techniques revolve around the idea of finding
the computation in $f$ that depends only on $x$, and performing them
once and for all to use.  Similarly, crucial compiler optimization
techniques such as loop hoisting and common subexpression elimination
identify common computations across all invocations of $f$ and perform
them once and for all at the beginning of the computation.  A
particular application area where such algorithm-design and code
optimization techniques are widely applied include real-time computer
graphics.  There, to achieve high-performance, modern graphics
architectures requires writing such programs.  Programmers therefore
manually separate code into distinct {\em passes} based on their
required frequency of execution.  For example, graphics computations
are expressed as a sequence of passes where each pass may perform a
computation at increasingly finer grain, e.g., per object, per region
of the screen, and per pixel, where later computations use the results
from the prior computations.

%
While performant, programming with manually separated passes results
in complex code where invariants must hold across different passes,
and local changes for one pass may require changes to other passes to
``plumb'' inputs from all other frequencies\,\cite{Foley:2011}.
%
In other words, manual pass separation breaks key programming
abstractions such as composition and modularity.
%
Graphics researchers therefore have suggested using explicitly staged
programming languages\,\cite{Proudfoot:2001,Foley:2011,He:2014},
deferring to the compiler the tedious task of separating the code into
passes. However, all prior efforts have been limited to simple
languages that do not support recursion or first-class functions.

Programming-languages researchers have studied similar problems. In
the 1980s, J{\o}rring and Scherlis identified {\em frequency
  reduction} or {\em precomputation}~\cite{JS86-staging} as the common
mechanisms for evaluating efficiently staged computations.  The idea
behind frequency reduction is to identify computations that are
performed multiple times and hoist them so that they can be performed
once and used later as needed.  The idea behind precomputation is to
identify computations that can be performed earlier---for example, at
compile time if their inputs are known statically---and perform them
at that earlier time.

There has been much work on language-based techniques for performing
precomputation. Partial evaluation~\cite{} achieves precomputation by
specializing functions for specific values.  Going back to our
example, we can specialize $f$ at $[1,2, \ldots, n]$, written
$f_{[1,2, \ldots, n]}$, such that $f(x,y) = f_{[1,2, \ldots, n]}(y)$
and specialize $F$ for the same input as
\begin{lstlisting}
fun @$F_{[1,2, \ldots, n]}(y_0, y_1 \ldots y_{m-1})$@ = 
  @$(f_{[1,2, \ldots, n]}(y_0);  f_{0}(y_1)  \ldots f_{0}(y_{m-1}))$@.
\end{lstlisting}
%
Such a partial evaluation typically proceed by performing a
binding-time analysis to determine the static and dynamic parts of $f$
and $F$ and specializes them based on the results of the binding-time
analysis. A closely related technique, {\em metaprogramming}, enables
the programmer to write programs that generate specialized code.
Metaprogramming enables fine-grained control over specialization by
requiring staging annotations that make explicit the stage of each
computation.

While previous work shows that precomputation can be performed
essentially automatically (via partial evaluation or meta
programming), frequency reduction has been less explored.  In their
original paper, J{\o}rring and Scherlis proposed {\em pass separation}
as a manual method for {\em splitting} a program into multiple passes.
%
Applied to our example, pass separation would split the function $f$
into $f_1$ and $f_2$ such that $f(x,y) = f_2(f_1(x),y)$.  We can then
rewrite our example as a multi-pass program:
%
\begin{lstlisting}
fun @$F(x, y_0, y_1 \ldots y_{m-1})$@ = 
  let @$fx = f_1(x)$@
  in @$(f_2(fx, y_0);  f_2(fx, y_1)  \ldots f_2(fx, y_{m-1}))$@.
\end{lstlisting}
%
The key difference between partial evaluation (meta programming) and
pass separation is that the latter does not require the first
argument~$x$: it works with all arguments, whereas the partially
evaluated program is specialized for a particular value of $x$ ($x =
[1,2, \ldots, n]$ in the example). Specifically, if desired, partial
evaluation can be applied subsequently in our example, by partially
evaluating function~$f_2$ on $f_1(x)$.

However, unlike for precomputation, there is no available technique
for pass separation except for very simple languages
\cite{knoblock96,Proudfoot:2001,Foley:2011,He:2014}.  Notably omitted
features include recursion and first-class functions.  We are thus
interested in the problem of splitting automatically a functional
language.
%

As our starting point, we take the modal language \lang\ with explicit
staging in the style of Davies~\cite{davies96}, where the ``circle''
modality denotes computation in a later stage, and present a staged
operational semantics similar for \lang\ (\ref{sec:stagedsemantics}).
We then present a splitting algorithm that stratifies a \lang\ program
into a two mono-staged programs, one for the first stage and for the
second stage (\ref{sec:splitting}).  Since each program is monostaged,
they are expressed in conventional functional languages.

\begin{comment}
\section{Introduction}

Such computations, where we apply a multivariate function to arguments
where one of the argument remains the same are common.  For example,
many algorithm-design techniques revolve around the idea of finding
the computation in $f$ that depends only on $x$, and performing them
once and for all to use.  Similarly, crucial compiler optimization
techniques such as loop hoisting and common subexpression elimination
identify common computations across all invocations of $f$ and perform
them once and for all at the beginning of the computation.  A
particular application area where such algorithm-design and code
optimization techniques are widely applied include real-time computer
graphics.  There, to achieve high-performance, programmers manually
separate code into distinct {\em passes} based on their required
frequency of execution.  For example, graphics computations are often
expressed as a sequence of passes where each pass may perform a
computation in increasingly finer grain, e.g., per object, per region
of the screen, and per pixel, where later computations use the results
from the prior computations.

%
While performant, this programming style of programming with manually
separated passes results in complex code where invariants must hold
across different passes, and local changes for one pass may require
changes to other passes to ``plumb'' inputs from all other
frequencies\,\cite{Foley:2011}.
%
In other words, manual pass separation breaks key programming
abstractions such as composition and modularity that we hold dear.
%
Graphics researchers therefore have suggested writing graphics
programs in explicitly staged programming
languages\,\cite{Proudfoot:2001,Foley:2011,He:2014}, deferring the
tedious task of separating the code into passes as needed for
performance to the compiler. However, all prior efforts have been
limited to simple languages that do not support recursion or
first-class functions.

Motivated by the crucial need for improved abstractions for performant
multi-stage programs, we investigate in this paper, the problem of
automatically splitting a higher-order program into a collection of
programs each of which gets executed in a single pass.
%
This is not a new problem.
%
Common theme in this problem domain is what in the 1980s J{\o}rring
and Scherlis called {\em frequency reduction} or {\em
  precomputation}~\cite{JS86-staging}.  The idea behind frequency
reduction is to identify computations that are performed multiple
times and hoist them so that they can be performed once and used later
as needed.  The idea behind precomputation is to identify computations
that can be performed earlier---for example, at compile time if their
inputs are known statically---and perform them at that earlier time.

@@

Partial evaluation achieves precomputation by distinguishing between
{\em static} and {\em dynamic} stages, and performing all static
computations first, followed by dynamic ones.  For example, suppose
that we wish to compute $F(x, y_0 \ldots y_{m-1})$ for a statically
known $x = [1,2, \ldots, n]$.  We can derive a specialization of $f$
at $[1,2, \ldots, n]$, written $f_{[1,2, \ldots, n]}$, such that
$f(x,y) = f_{[1,2, \ldots, n]}(y)$ and specialize $F$ for $x=[~]$ as
\begin{lstlisting}
fun @$F_{[1,2, \ldots, n]}(y_0, y_1 \ldots y_{m-1})$@ = 
  @$(f_{[1,2, \ldots, n]}(y_0);  f_{0}(y_1)  \ldots f_{0}(y_{m-1}))$@.
\end{lstlisting}
%
Such a partial evaluation typically proceed by performing a
binding-time analysis to determine the static and dynamic parts of $f$
and $F$ and specializes them based on the results of the binding-time
analysis.

@@

The third and final approach, called {\em pass separation} or {\em
  data specialization}, assumes much less information: instead of
requiring the static data, it requires just the knowledge of which
computations are static and dynamic and uses this knowledge to
generate two stage computations. 
%
Applied to our example, pass separation would {\em split} the function
$f$ into $f_s$ and $f_d$, corresponding to the static and dynamic
passes, such that $f(x,y) = f_d(f_s(x),y)$.  We can then rewrite our
example as
%
\begin{lstlisting}
fun @$F(x, y_0, y_1 \ldots y_{m-1})$@ = 
let @$fx = f_s(x)$@
in @$(f_d(fx, y_0);  f_d(fx, y_1)  \ldots f_d(fx, y_{m-1}))$@.
\end{lstlisting}
We refer to such a program as a {\em multi-pass} program.

The key difference with partial evaluated program and the stratified
program is that the latter has lost no generality: it works with all
arguments, whereas the partially evaluated program is specialized for
a particular value of $x$ ($x = [1,2, \ldots, n]$ in the example).
Thus, the pass separation can be applied without leading to any loss
of generality.  In fact, this is quite natural---in the terminology of
J{\o}rring and Scherlis, pass separation is a frequency reduction,
whereas as partial evaluation and metaprogramming are precomputation
techniques. Consequently, if desired, partial evaluation can be
orthogonally applied to a multi-pass program.
%
The crucial benefit of pass-separation is efficiency: separating the
passes in the program factors out frequently used the parts of
computation, computes them first, and then reuse them later instead of
computing them again. This is precisely the reason for why it is
employed broadly in graphics software even though it results in
code that is difficult to design, implement, and maintain.

Thus, we are interested in the question of whether it would be
possible to obtain the best of the both worlds, by automatically
pass-separating programs.  Such automatic stratification has been
shown to be challenging. The first successful approach that we know of
is by Knoblock and Ruf\,\cite{knoblock96}, which, like our work, was
inspired by computer graphics applications.  While practically
effective, Knoblock and Ruf consider only a very simple language and
are not able to stratify recursive programs.  Recent
efforts\,\cite{Proudfoot:2001,Foley:2011,He:2014} ....  but still
cannot stratify recursive programs.

In this paper we make significant progress on the problem of automatic
stratification of programs expressed in a rich higher-order functional
language.  As our starting point, we take the modal language \lang\
with explicit staging in the style of Davies~\cite{davies96}, where
the ``circle'' modality denotes computation in a later stage, and
present a staged operational semantics similar for \lang\
(\ref{sec:stagedsemantics}).  We then present a splitting algorithm
that stratifies a \lang\ program into a two mono-staged programs, one
for the first stage and for the second stage (\ref{sec:splitting}).
Since each program is monostaged, they are expressed in conventional
functional languages.

@@

We start in \ref{sec:example} with an extended example of splitting
\texttt{quickselect}.
In \ref{sec:semantics}, we describe our staged language \lang, including its
type system and operational semantics.
In \ref{sec:splitting,sec:implementation}, we describe our stage-splitting
algorithm for performing data specialization.
Finally, in \ref{sec:examples}, we show how our stage-splitting algorithm
transforms a variety of other programs.
\end{comment}
